 Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation

### 2021-05-31 analysis/paper/shceof.Rmd
Mueve todo a wrap por oración.

Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis). Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two Pacific–South American Patterns (PSA) [@mo2001]. The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector. The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula. These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.

EOFs are more flexible than Fourier descomposition modes in the sense that they can characterise \textcolor{red}{Esto hay que exresarlo más claramente}. Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves. Besides, a single EOF can also represent a mixture of two or more physical modes. The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs. @irving2016 characterises the Pacific–South American Pattern in a Fourier framework. Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator. On this new projection, they could identify the PSA using Fourier decomposition.  
Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences. The Marshal SAM Index [@marshall2003], for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following @gong1999. @raphael2004 used a similar methodology to describe the wave 3 in the Southern Hemisphere. Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean. This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field. Similarly, @hobbs2010 propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere. These methods are grounded on other methods to identify the centers of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods. The use of data of only specific regions or locations makes them suitable for easy computation and for extending into less quality data from the past, but also makes them not very robust and could not capture non-stationary behaviours of the patterns.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019. Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends. Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK [@fishpack] and horizontal wave activity fluxes was computed following @plumb1985. We usded Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree\ resolution in latitude and longitude. This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

In traditional EOF analysis zonal waves appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase [@horel1984]. For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa \textcolor{red}{Creo que no aclaras ni acá, ni para los complejos, cuál sería el dominio para el cuál estás calculando el EOF) La barra de colores quedaría mejor si le agregas un contorno que la encierre.}. It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength. A similar EOF structure can be seen in 200 hPa (not shown). Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

@@

  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```
For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`. This also nearly minimises the correlation with the imaginary part.

@@

We performed linear regressions to quantify the association between the cEOFs and other variables. For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part. We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate [@benjamini1995;@wilks2016] to avoid misleading results from the high number of regressions [@walker1914;@katz1991].

@@

We performed all analysis in this paper using the R programming language [@rcoreteam2020], using data.table [@dowle2020] and metR [@campitelli2020] packages. All graphics are made using ggplot2 [@wickham2009]. We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020]. The paper was rendered using knitr and rmarkdown [@xie2015;@allaire2019].

@@

To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, compueted jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2). The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).  In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the temporal cEOF. The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive).

The first cEOF (Fig.\ \@ref(fig:eof-naive) column\ a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes. At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase. The second cEOF (Fig.\ \@ref(fig:eof-naive) column\ b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales. In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector. This modulated amplitude is especially apparent at 200 hPa. There is no apparent phase shift, suggesting a rather barotropic structure.

The temporal variability for both components of the cEOF1 have non zero mean (Fig.\ \@ref(fig:ceofs-2)). This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean. As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field. There are no significant simultanous correlation between the time series. Both cEOFs show year-to-year variability but show no evidence of decadal variabilty. The relatively short timespan (`r diff(range(unique(year(era5$time))))` years) hampers our habilty to detect longer trends.

@@

Therefore, we extend the cEOFs time series using the preliminary ERA5 back extension going back to 1950 by projecting the older SON geopogential height zonal anomaly fields onto the spatial patterns shown in Figure\ \@ref(fig:ceofs-1). There is a significant upward trend in the real component of cEOF1 (Fig.\ \@ref(fig:extended-series)a.1, p-value\ `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2. The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).

@@

The spatial patterns shown in Figure\ \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series. To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields.

@@

Figure\ \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1. At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure\ \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica. However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

The regression patterns at 200 hPa (Figure\ \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation. It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure\ \@ref(fig:eof1-regr-gh)b.1). The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.

With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.\ \@ref(fig:ceofs-1)), particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa. Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere, they are associated with slightly off-centre monopoles.

(ref:eof2-regr-gh-cap) Same as Figure\ \@ref(fig:eof1-regr-gh) but for the second cEOF.

@@

Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height anomalies upon the cEOF2.  Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure\ \@ref(fig:ceofs-1). Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans. cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001]. Comparing the location of the positive anomaly near 90ºW in column b of Figure\ \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.

@@

(ref:eof1-regr-t-cap) Same as Figure\ \@ref(fig:eof1-regr-gh) but for air temperature.

@@

The relation between cEOFs and temperature was also evaluated. Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1. In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.\ \@ref(fig:eof1-regr-t) column 1). On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE. The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.\ \@ref(fig:eof1-regr-gh)).

Figure \ \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficientes on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`. \textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando. Lo podes hacer en los métodos y ya queda para todo el texto}. Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa  with a change in sign above 10 hPa. Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.\ \@ref(fig:t-vertical))). Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere . This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere. In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979;@wirth1993;@smith1995].

@@

The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig. \ \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1. Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993, \textcolor{red}{¿Alguna referencia más nueva?}]. Thus, the Real cEOF1 regression pattern (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.

@@

There is a close spatial relationship between amplitudes and phases of the cEOF1 and TOC planetary wave 1 between `r knitr::combine_words(LatLabel(o3wave_lats))`. The correlation between the amplitude of both indices is `r wave1_cor[with == "amplitude"]$text` the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`. As expected from the location of regressions in Figure\ \@ref(fig:o3-regr), the Real cEOF1 drives the relationship of amplitudes and the Imaginary cEOF1 drives the relationship of phases between the timeseries (not shown).

@@

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021. Thick lines represent estimates with p-value < 0.01 corrected for False Detection Rate [@benjamini1995].

@@

To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig.\ \@ref(fig:sam-eof-vertical)). The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere. It is not correlated with the Real cEOF2 at any level.

The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa. In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM. These correlations are consistent with the regression patterns of geopotential height in Figure\ \@ref(fig:eof1-regr-gh).

@@

In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered. The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure\ \@ref(fig:sam-eof-vertical).b2). Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.

<!-- As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ \@ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed. The resulting pattern is an nearly zonally symmetrical annular mode. This mode is highly correlated with SAM (`r eof_filtered_cor[term == "full"]$text`), but only with the symmetrical part (`r eof_filtered_cor[term == "sym"]$text`). Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.   -->

@@

To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index. To this synthetic dataset (\textcolor{red}{No me suena mucho la palabra "synthetic", ¿se usa en este contexto?}) we then apply traditional EOF. By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure\ \@ref(fig:fake-eof) row a). While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so it needs two EOFs to be represented correctly. Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.

@@

## Code availability {-}

>>>>

Another approach to characterising the Southern Hemisphere circulation is by using Empirical Orthogonal Functions (EOF, also known as Principal Component Analysis).
Within the EOF framework, the Southern Annular Mode (SAM) appears as the leading mode of variability of the tropospheric Southern Hemisphere circulation [@fogt2020] followed by the two Pacific--South American Patterns (PSA) [@mo2001].
The SAM represents a relatively zonally symmetric pattern of alternating low pressures in polar latitude and a ring of high pressures in high latitudes with an embedded wave 3 pattern that is more prominent in the Pacific sector.
The PSA1 and PSA2 describe the two out-of-phase sides of a meridionally propagating wave train that originates in the Eastern equatorial Pacific and travels towards the South Atlantic following a great-circle arch along the Antarctic Peninsula.
These patterns are derived by applying EOF to temporal anomalies, but @raphael2003 applied EOF methods specifically to zonal anomalies.

EOFs are more flexible than Fourier descomposition modes in the sense that they can characterise \textcolor{red}{Esto hay que exresarlo más claramente}.
Nonetheless, they are restricted to standing oscillation modes and could not represent properly propagating or phase-varying modes such as zonal waves.
Besides, a single EOF can also represent a mixture of two or more physical modes.
The inability of a single EOF to describe travelling waves is what forces the PSA pattern to be described by two EOFs.
@irving2016 characterises the Pacific--South American Pattern in a Fourier framework.
Being a meridionally propagating mode, the PSA cannot be correctly characterised by Fourier decomposition at each latitude circle, so they reprojected meridional wind fields so that the path of the PSA laid on the equator.
On this new projection, they could identify the PSA using Fourier decomposition.

Finally, another, hybrid, methodology commonly used consist on identifying particular features of interest and creating indices using simple methods such as averages and differences.
The Marshal SAM Index [@marshall2003], for example, tries to characterise the SAM by the average difference of sea level pressure between 40ºS and 65ºS, following @gong1999.
@raphael2004 used a similar methodology to describe the wave 3 in the Southern Hemisphere.
Instead of using Fourier to compute the amplitude and phase of the zonal wave, they averaged standardised geopotential height anomalies in three points representing the location of the ridges of the zonal wave 3 of the climatological mean.
This is equivalent to an index of the amplitude of the zonal wave 3 that projects into the phase of the zonal wave 3 of the mean field.
Similarly, @hobbs2010 propose that the zonally asymmetric circulation in the Southern Hemisphere can be described by the strength and location of two anticyclones located in the sub-Antarctic western hemisphere.
These methods are grounded on other methods to identify the centers of action for the described phenomena and can be useful to characterise features that are not readily apparent with Fourier or EOF methods.
The use of data of only specific regions or locations makes them suitable for easy computation and for extending into less quality data from the past, but also makes them not very robust and could not capture non-stationary behaviours of the patterns.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019.
Most of our analysis was restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends.
Streamfunction was derived from ERA5 vorticity at 200 hPa using the FORTRAN subroutine FISHPACK [@fishpack] and horizontal wave activity fluxes was computed following @plumb1985.
We usded Sea Surface Temperature (SST) monthly fields from Extended Reconstructed Sea Surface Temperature (ERSST) v5 [@huang2017] and precipitation monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree resolution in latitude and longitude.
This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

In traditional EOF analysis zonal waves appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase [@horel1984].
For instance, Figure \@ref(fig:eof-naive) shows the leading 4 EOFs of SON geopotential height zonal anomalies at 50 hPa \textcolor{red}{Creo que no aclaras ni acá, ni para los complejos, cuál sería el dominio para el cuál estás calculando el EOF) La barra de colores quedaría mejor si le agregas un contorno que la encierre.}.
It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
A similar EOF structure can be seen in 200 hPa (not shown).
Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

@@

  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between `r knitr::combine_words(LatLabel(o3wave_lats))`.
This also nearly minimises the correlation with the imaginary part.

@@

We performed linear regressions to quantify the association between the cEOFs and other variables.
For each cEOF, we create regression maps by computing fitting a multiple linear model involving both the real and the imaginary part.
We evaluated statistical significance using a two-sided t-test and, in the case of regression maps, p-values were adjusted by controlling for the False Discovery Rate [@benjamini1995; @wilks2016] to avoid misleading results from the high number of regressions [@walker1914; @katz1991].

@@

We performed all analysis in this paper using the R programming language [@rcoreteam2020], using data.table [@dowle2020] and metR [@campitelli2020] packages.
All graphics are made using ggplot2 [@wickham2009].
We downloaded data from reanalysis using the ecmwfr package [@hufkens2020] and indices of the ENSO with the rsoi package [@albers2020].
The paper was rendered using knitr and rmarkdown [@xie2015; @allaire2019].

@@

To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, compueted jointly at both levels, are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2).
The first mode (cEOF1) explains 82% of the variance, while the second mode (cEOF2) explains a smaller fraction (7%).
In the spatial patterns (Fig. \@ref(fig:ceofs-1)), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the temporal cEOF.
The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure \@ref(fig:eof-naive).

The first cEOF (Fig. \@ref(fig:eof-naive) column a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
The second cEOF (Fig. \@ref(fig:eof-naive) column b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, its dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
This modulated amplitude is especially apparent at 200 hPa.
There is no apparent phase shift, suggesting a rather barotropic structure.

The temporal variability for both components of the cEOF1 have non zero mean (Fig. \@ref(fig:ceofs-2)).
This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean.
As a consequence, the variability associated with the cEOF1 includes variability that projects onto the mean zonally anomalous field.
There are no significant simultanous correlation between the time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variabilty.
The relatively short timespan (`r diff(range(unique(year(era5$time))))` years) hampers our habilty to detect longer trends.

@@

Therefore, we extend the cEOFs time series using the preliminary ERA5 back extension going back to 1950 by projecting the older SON geopogential height zonal anomaly fields onto the spatial patterns shown in Figure \@ref(fig:ceofs-1).
There is a significant upward trend in the real component of cEOF1 (Fig. \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & cEOF == "cEOF1"]$p_report`) and no significant trend in any of the complex components of cEOF2.
The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).

@@

The spatial patterns shown in Figure \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series.
To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields.

@@

Figure \@ref(fig:eof1-regr-gh) shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º off phase.
Their phases coincide with the ones shown in Figure \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica.
However, the Real cEOF1 pattern is substantially altered by the zonally symmetric circulation.
Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

The regression patterns at 200 hPa (Figure \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation.
It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure \@ref(fig:eof1-regr-gh)b.1).
The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.

With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig. \@ref(fig:ceofs-1)), particularly at 200 hPa.
Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
While in the troposphere, they are associated with slightly off-centre monopoles.

(ref:eof2-regr-gh-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for the second cEOF.

@@

Figure \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height anomalies upon the cEOF2. Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure \@ref(fig:ceofs-1). Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans. cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001].
Comparing the location of the positive anomaly near 90ºW in column b of Figure \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.

@@

(ref:eof1-regr-t-cap) Same as Figure \@ref(fig:eof1-regr-gh) but for air temperature.

@@

The relation between cEOFs and temperature was also evaluated.
Figure \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1.
In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig. \@ref(fig:eof1-regr-t) column 1).
On the other hand, the regression maps on the Imaginary cEOF1 show a more clear wave 1 pattern with its maximum around 60ºE.
The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig. \@ref(fig:eof1-regr-gh)).

Figure  \@ref(fig:t-vertical) shows the vertical distribution of the regression coefficientes on cEOF1 from zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between `r knitr::combine_words(LatLabel(o3wave_lats))`.
\textcolor{red}{Deberias aclarar qué significa la terminología que estas utilizando. Lo podes hacer en los métodos y ya queda para todo el texto}.
Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa.
Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig. \@ref(fig:t-vertical))). Therefore, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere . This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere. In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979; @wirth1993; @smith1995].

@@

The regression maps of cEOF1 with fields of Total Ozone Column (TOC) (Fig.  \@ref(fig:o3-regr)) show zonal wave 1 pattens in TOC associated with both phases of cEOF1. Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993, \textcolor{red}{¿Alguna referencia más nueva?}].
Thus, the Real cEOF1 regression pattern (Figure \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary cEOF1 is shifted by 90º.

@@

There is a close spatial relationship between amplitudes and phases of the cEOF1 and TOC planetary wave 1 between `r knitr::combine_words(LatLabel(o3wave_lats))`.
The correlation between the amplitude of both indices is `r wave1_cor[with == "amplitude"]$text` the overall correlation between the zonal wave 1 of ozone and cEOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`.
As expected from the location of regressions in Figure \@ref(fig:o3-regr), the Real cEOF1 drives the relationship of amplitudes and the Imaginary cEOF1 drives the relationship of phases between the timeseries (not shown).

@@

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021. Thick lines represent estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].

@@

To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig. \@ref(fig:sam-eof-vertical)).
The SAM index is moderately correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
It is not correlated with the Real cEOF2 at any level.

The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
These correlations are consistent with the regression patterns of geopotential height in Figure \@ref(fig:eof1-regr-gh).

@@

In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
The Imaginary cEOF2 and the A-SAM share up to `r scales::percent(max_r2[cEOF == "cEOF2"]$estimate^2, 2)` of their variance, reached at `r max_r2[cEOF == "cEOF2"]$lev` hPa (Figure \@ref(fig:sam-eof-vertical).b2).
Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.

<!-- As a further illustration \textcolor{red}{(¿De qué?)}, Figure\ @ref(fig:eof-filtered) shows the spatial pattern of the leading EOF of 200 hPa geopotential height when the variability associated with cEOF2 is removed. The resulting pattern is an nearly zonally symmetrical annular mode. This mode is highly correlated with SAM (`r eof_filtered_cor[term == "full"]$text`), but only with the symmetrical part (`r eof_filtered_cor[term == "sym"]$text`). Note that the usual definition of the PSA modes as the second and third EOFs creates modes orthogonal to the SAM (defined as the first EOF) and thus impedes this kind of filtering.   -->

@@

To test the hypothesis that a PSA-like feature can appear as statistic contamination of an otherwise zonally symmetrical and uncorrelated SAM, we create a synthetic dataset of `r uniqueN(fake_sam$time)` fields by reconstructing the SON 200 hPa geopotential height field explained by the cEOF2 and adding a perfectly zonally symmetric SAM-like structure with random sign and value sampled with replacement from the real SAM index.
To this synthetic dataset (\textcolor{red}{No me suena mucho la palabra "synthetic", ¿se usa en este contexto?}) we then apply traditional EOF.
By construction this dataset contains two independent components: a zonally symmetric standing oscillation and a PSA-like of varying amplitude and phase (Figure \@ref(fig:fake-eof) row a).
While the former is a standing oscillation which could be represented by a single EOF, the latter is a wave which can change phase, so it needs two EOFs to be represented correctly.
Thus, the full variance of this synthetic dataset can be perfectly explained by three EOFs.

@@

## Code availability {.unnumbered}
